Tail Flexibility in the Degrees of Preferential Attachment Networks

### 2025-04-11 paper.qmd
Updating draft

---
title: Tail Flexibility in the Degrees of Preferential
 Attachment Networks
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-basic
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    number-sections: true
author:
  - name: Thomas William Boughen
    affiliations:
      - name: Newcastle University
        id: 1
        department: School of Mathematics, Statistics and Physics
  - name: Clement Lee
    affiliations:
      - ref: 1
  - name: Vianey Palacios Ramirez
    affiliations:
      - ref: 1
abstract: |
  Devising the underlying generating mechanism of a real-life network is difficult as, more often than not, only its snapshots are available, but not its full evolution. One candidate for the generating mechanism is preferential attachment which, in its simplest form, results in a degree distribution that follows the power law. Consequently, the growth of real-life networks that roughly display such power-law behaviour is commonly modelled by preferential attachment. However, the validity of the power law has been challenged by the presence of alternatives with comparable performance, as well as the recent findings that the right tail of the degree distribution is often lighter than implied by the body, whilst still being heavy. In this paper, we study a modified version of the model with a flexible preference function that allows super/sub-linear behaviour whilst also guaranteeing that the limiting degree distribution has a heavy tail. We relate the distributions tail heaviness directly to the model parameters, allowing direct inference of the parameters from the degree distribution alone.
keywords: [networks, extremes]
bibliography: refs.bib  
---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(paletteer)
library(gridExtra)
library(latex2exp)
library(ggpubr)
```

@@

The procedure for recovering the parameters begins with simulating a network from the model with $N=100,000$ vertices and $m=1$ given some set of parameters $\pmb\theta = (\alpha, \beta, \varepsilon, k_0)$, obtaining the degree counts in the form of a vector of degrees $\pmb x = (1,2,\ldots,M)$ and the number of vertices with those degrees $\pmb n$ where $M$ is the maximum degree.

Using these and the p.m.f derived in [@rudas07], the likelihood given that $x_i \ge l$ for all $i =1,2,\ldots,M$ is:

@@

to obtain a posterior distribution that can then be used in an adaptive Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm  to obtain posterior samples. The results of this inference are shown in @fig-rec1 and @fig-rec2. For these simulated networks $l=2$.

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="80%"}
#| fig-width: 8
#| fig-height: 8
#| fig-cap: Posterior estimates of survival function for data simulated from the proposed model with various combinations of ($\alpha$,$\beta$,$\varepsilon$) and $k_0=20$.
#| label: fig-rec1
surfit_list = list()
pars = readRDS('recovery_pars.rds')
recovery_list = readRDS('recovery_dat.rds')
selected = c(1,27,14,35)
# selected = 1:36
for(k in 1:length(selected)){
  j=selected[k]
  x = sort(unique(recovery_list[[j]]$degs))
  ls = c()

  y_975 = recovery_list[[j]]$mcmc$surv$CI[2,]
  y_025 = recovery_list[[j]]$mcmc$surv$CI[1,]
  y_50 = recovery_list[[j]]$mcmc$surv$est

  surfit_list[[k]] = ggplot() + geom_point(data=twbfn::deg_surv(recovery_list[[j]]$degs), aes(x=degree, y=surv)) +
    geom_line(data=NULL, aes(x=!!x, y=!!y_975), colour = 'red', linetype='dashed')+
    geom_line(data=NULL, aes(x=!!x, y=!!y_50),colour='red')+
    geom_line(data=NULL, aes(x=!!x, y=!!y_025), colour = 'red', linetype='dashed')+
    scale_x_log10(limits=c(1,1e5))  +scale_y_log10(limits = c(1/length(recovery_list[[j]]$degs),1))+theme_bw() + theme(aspect.ratio = 0.66,axis.title.y=element_blank(),                                                                   axis.text.y=element_blank(),                                                                     axis.ticks.y=element_blank()) + xlab('')+ylab('') +
    ggtitle(TeX(paste0('$\\alpha = $',pars$a[j],
                       ', $\\epsilon = $',pars$eps[j],
                       ', $\\beta = $',pars$b[j])))
}

fig = ggpubr::ggarrange(plotlist = surfit_list,
          nrow=round(sqrt(length(surfit_list)),0),ncol=round(sqrt(length(surfit_list)),0),
          label.x='Degree', label.y = 'Survival')

ggpubr::annotate_figure(fig, bottom='Degree', left='Survival')

@@

Turning now to real data, the goal is to fit the model to the degree distributions of real networks from various sources. [describe networks being used].

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="80%"}
#| fig-width: 9
#| fig-height: 6
#| fig-cap: Posterior estimates (solid red) of survival for several real data sets and their 95% credible intervals (dotted red)
#| label: fig-real1

plt1 = readRDS('real_survs.rds')

@@

@fig-real1 displays the posterior estimates of the conditional survival $F(k|k\ge2)$ for several data sets, for many of the data sets the fit is quite good  whilst for other the fit is very bad. This may indicate that the data sets without a good fit are unlikely to have come from a preferential attachment model. However for those with a good fit, it is possible to obtain information about the underlying PA mechanism assuming that the corresponding networks did indeed grow in that fashion.

To demonstrate that the model is capturing the tail behaviour of the degree distributions @fig-gp_comp shows the tail heaviness based on samples from the MCMC for each network as well as the tail heaviness as estimated by fitting the GPD to the degrees for various thresholds.

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="80%"}
#| fig-width: 9
#| fig-height: 6
#| fig-cap: Posterior samples of shape parameter from model (blue) and GPD shape parameter (red) at various thresholds
#| label: fig-gp_comp

plt2 = readRDS('gp_plt.rds')

>>>>

---
title: Tail Flexibility in the Degrees of Preferential
 Attachment Networks
format:
  nature-pdf:
    fig-pos: 'H'
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-basic
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    number-sections: true
author:
  - name: Thomas William Boughen
    affiliations:
      - name: Newcastle University
        id: 1
        department: School of Mathematics, Statistics and Physics
  - name: Clement Lee
    affiliations:
      - ref: 1
  - name: Vianey Palacios Ramirez
    affiliations:
      - ref: 1
abstract: |
  Devising the underlying generating mechanism of a real-life network is difficult as, more often than not, only its snapshots are available, but not its full evolution. One candidate for the generating mechanism is preferential attachment which, in its simplest form, results in a degree distribution that follows the power law. Consequently, the growth of real-life networks that roughly display such power-law behaviour is commonly modelled by preferential attachment. However, the validity of the power law has been challenged by the presence of alternatives with comparable performance, as well as the recent findings that the right tail of the degree distribution is often lighter than implied by the body, whilst still being heavy. In this paper, we study a modified version of the model with a flexible preference function that allows super/sub-linear behaviour whilst also guaranteeing that the limiting degree distribution has a heavy tail. We relate the distributions tail heaviness directly to the model parameters, allowing direct inference of the parameters from the degree distribution alone.
keywords: [networks, extremes]
bibliography: refs.bib  
---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(paletteer)
library(gridExtra)
library(latex2exp)
library(ggpubr)
library(Rcpp)
library(RcppArmadillo)
```

@@

The procedure for recovering the parameters begins with simulating a network from the model with $N=100,000$ vertices and $m=1$ given some set of parameters $\pmb\theta = (\alpha, \beta, \varepsilon, k_0)$, obtaining the degree counts in the form of a vector of degrees $\pmb x = (1,2,\ldots,M)$ and the number of vertices with those degrees $\pmb n$ where $M$ is the maximum degree. However, when it comes to fitting this model to real data in the next section, small degrees are extremely influential and will often cause the model to have a bad fit even though the rest of the data looks similar to that of the simulations; truncating the data above some value, say $l<k_0$, should allow the model to be more effectively fitted. This accounts in some way for the fact that this model only applies to trees, and could not possibly capture the behaviour for nodes with small degrees. Using the p.m.f derived in [@rudas07], the likelihood given that $x_i \ge l$ for all $i =1,2,\ldots,M$ is:

@@

to obtain a posterior distribution that can then be used in an adaptive Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm  to obtain posterior samples. The results of this inference are shown in @fig-rec1 and @fig-rec2. For these simulated networks $l=0$.

```{r, echo=FALSE, warning=FALSE, cache=FALSE, out.width="80%"}
#| fig-width: 8
#| fig-height: 8
#| fig-cap: Posterior estimates of survival function for data simulated from the proposed model with various combinations of ($\alpha$,$\beta$,$\varepsilon$) and $k_0=20$.
#| label: fig-rec1
# surfit_list = list()
# pars = readRDS('recovery_pars.rds')
# recovery_list = readRDS('recovery_dat.rds')
# selected = c(1,27,14,35)
# # selected = 1:36
# for(k in 1:length(selected)){
#   j=selected[k]
#   x = sort(unique(recovery_list[[j]]$degs))
#   ls = c()
#   
#   y_975 = recovery_list[[j]]$mcmc$surv$CI[2,]
#   y_025 = recovery_list[[j]]$mcmc$surv$CI[1,]
#   y_50 = recovery_list[[j]]$mcmc$surv$est
#   
#   surfit_list[[k]] = ggplot() + geom_point(data=twbfn::deg_surv(recovery_list[[j]]$degs), aes(x=degree, y=surv)) +
#     geom_line(data=NULL, aes(x=!!x, y=!!y_975), colour = 'red', linetype='dashed')+
#     geom_line(data=NULL, aes(x=!!x, y=!!y_50),colour='red')+
#     geom_line(data=NULL, aes(x=!!x, y=!!y_025), colour = 'red', linetype='dashed')+
#     scale_x_log10(limits=c(1,1e5))  +scale_y_log10(limits = c(1/length(recovery_list[[j]]$degs),1))+theme_bw() + theme(aspect.ratio = 0.66,axis.title.y=element_blank(),                                                                   axis.text.y=element_blank(),                                                                     axis.ticks.y=element_blank()) + xlab('')+ylab('') +
#     ggtitle(TeX(paste0('$\\alpha = $',pars$a[j],
#                        ', $\\epsilon = $',pars$eps[j],
#                        ', $\\beta = $',pars$b[j])))
# }
# 
# fig = ggpubr::ggarrange(plotlist = surfit_list,
#           nrow=round(sqrt(length(surfit_list)),0),ncol=round(sqrt(length(surfit_list)),0),
#           label.x='Degree', label.y = 'Survival')
# 
# 
# ggpubr::annotate_figure(fig, bottom='Degree', left='Survival')

@@

Turning now to real data, the goal is to fit the model to the degree distributions of real networks from various sources. [describe networks being used]. Alongside fitting this model to the degree distributions, we then compare the fit to that of an mixture distribution that was used in [clement] and is similar to others that have been used to model degree distributions [others].

```{r, echo=FALSE, warning=FALSE, cache=FALSE, out.width="80%"}
#| fig-width: 9
#| fig-height: 9
#| fig-cap: Posterior estimates (solid red) of survival for several real data sets and their 95% credible intervals (dotted red)
#| label: fig-real1

plt1 = readRDS('comp_plot.rds')

@@

@fig-real1 displays the posterior estimates of the survival function for various data sets, obtained from fitting the GPA model and the Zipf-IGP mixture model. In most cases, the GPA model does not necessarily provide an improvement in fit when compared to the Zipf-IGP model but where the GPA model fits well we gain additional information about the preference function assuming that the network evolved according the the GPA scheme. @fig-shapes shows the posterior of the shape parameter $\xi$ obtained from the Zipf-IGP model alongside the posterior of the equivalent shape parameter $\beta/\lambda^*$ obtained from fitting the GPA model.

```{r, echo=FALSE, warning=FALSE, cache=FALSE, out.width="80%"}
#| fig-width: 9
#| fig-height: 9
#| fig-cap: Posterior estimates (solid red) of survival for several real data sets and their 95% credible intervals (dotted red)
#| label: fig-shapes

plt2 = readRDS('shape_plt.rds')

### 2025-04-11 paper.tex
Updating draft

\begin{figure}

@@

The procedure for recovering the parameters begins with simulating a
network from the model with \(N=100,000\) vertices and \(m=1\) given
some set of parameters
\(\pmb\theta = (\alpha, \beta, \varepsilon, k_0)\), obtaining the degree
counts in the form of a vector of degrees \(\pmb x = (1,2,\ldots,M)\)
and the number of vertices with those degrees \(\pmb n\) where \(M\) is
the maximum degree.

Using these and the p.m.f derived in \citep{rudas07}, the likelihood
given that \(x_i \ge l\) for all \(i =1,2,\ldots,M\) is:

@@

to obtain a posterior distribution that can then be used in an adaptive
Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm to obtain
posterior samples. The results of this inference are shown in
Figure~\ref{fig-rec1} and Figure~\ref{fig-rec2}. For these simulated
networks \(l=2\).

\begin{figure}

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-rec1-1.pdf}

\caption{\label{fig-rec1}Posterior estimates of survival function for
data simulated from the proposed model with various combinations of
(\(\alpha\),\(\beta\),\(\varepsilon\)) and \(k_0=20\).}

\begin{figure}

@@

Turning now to real data, the goal is to fit the model to the degree
distributions of real networks from various sources. {[}describe
networks being used{]}.

\begin{figure}

@@

Figure~\ref{fig-real1} displays the posterior estimates of the
conditional survival \(F(k|k\ge2)\) for several data sets, for many of
the data sets the fit is quite good whilst for other the fit is very
bad. This may indicate that the data sets without a good fit are
unlikely to have come from a preferential attachment model. However for
those with a good fit, it is possible to obtain information about the
underlying PA mechanism assuming that the corresponding networks did
indeed grow in that fashion.

To demonstrate that the model is capturing the tail behaviour of the
degree distributions Figure~\ref{fig-gp_comp} shows the tail heaviness
based on samples from the MCMC for each network as well as the tail
heaviness as estimated by fitting the GPD to the degrees for various
thresholds.

\begin{figure}

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-gp_comp-1.pdf}

@@

\caption{\label{fig-gp_comp}Posterior samples of shape parameter from
model (blue) and GPD shape parameter (red) at various thresholds}

>>>>

@@

The procedure for recovering the parameters begins with simulating a
network from the model with \(N=100,000\) vertices and \(m=1\) given
some set of parameters
\(\pmb\theta = (\alpha, \beta, \varepsilon, k_0)\), obtaining the degree
counts in the form of a vector of degrees \(\pmb x = (1,2,\ldots,M)\)
and the number of vertices with those degrees \(\pmb n\) where \(M\) is
the maximum degree. However, when it comes to fitting this model to real
data in the next section, small degrees are extremely influential and
will often cause the model to have a bad fit even though the rest of the
data looks similar to that of the simulations; truncating the data above
some value, say \(l<k_0\), should allow the model to be more effectively
fitted. This accounts in some way for the fact that this model only
applies to trees, and could not possibly capture the behaviour for nodes
with small degrees. Using the p.m.f derived in \citep{rudas07}, the
likelihood given that \(x_i \ge l\) for all \(i =1,2,\ldots,M\) is:

@@

to obtain a posterior distribution that can then be used in an adaptive
Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm to obtain
posterior samples. The results of this inference are shown in
\textbf{?@fig-rec1} and Figure~\ref{fig-rec2}. For these simulated
networks \(l=0\).

@@

Turning now to real data, the goal is to fit the model to the degree
distributions of real networks from various sources. {[}describe
networks being used{]}. Alongside fitting this model to the degree
distributions, we then compare the fit to that of an mixture
distribution that was used in {[}clement{]} and is similar to others
that have been used to model degree distributions {[}others{]}.

@@

Figure~\ref{fig-real1} displays the posterior estimates of the survival
function for various data sets, obtained from fitting the GPA model and
the Zipf-IGP mixture model. In most cases, the GPA model does not
necessarily provide an improvement in fit when compared to the Zipf-IGP
model but where the GPA model fits well we gain additional information
about the preference function assuming that the network evolved
according the the GPA scheme. Figure~\ref{fig-shapes} shows the
posterior of the shape parameter \(\xi\) obtained from the Zipf-IGP
model alongside the posterior of the equivalent shape parameter
\(\beta/\lambda^*\) obtained from fitting the GPA model.

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-shapes-1.pdf}

@@

\caption{\label{fig-shapes}Posterior estimates (solid red) of survival
for several real data sets and their 95\% credible intervals (dotted
red)}
